---
title: "100 - RoSA map figure 1"
author: "Neil Thompson"
date: "Last Updated: `r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
start_time <- Sys.time()
```

```{r}
library(raster)
library(sf)
library(tidyverse)
library(ggspatial)
library(ggsn)
library(cowplot)
dir.create("./outputs/100", recursive = TRUE, showWarnings = FALSE)
```

# Prepare spatial data

## Grab basemap and state lines

```{r}
nat.earth <- brick('geo-spatial/NE2_HR_LC_SR_W_DR/NE2_HR_LC_SR_W_DR.tif') #
state_prov <- st_read("geo-spatial/ne_10m_admin_1_states_provinces_lines/ne_10m_admin_1_states_provinces_lines.shp")
coastline <- st_read("geo-spatial/ne_10m_coastline/ne_10m_coastline.shp")
```

## Get river line data

Download shapefiles for each watershed using https://viewer.nationalmap.gov/basic/ or for California Lakes and Rivers (https://data.cnra.ca.gov/dataset/national-hydrography-dataset-nhd)

To save the hassle of the above downloads, we have read
those shapefiles into variables using the code (not evaluated
by default) and saved that into `stored_results/100/rivers.rda`
```{r, eval=FALSE}
cali_rivers <- st_read("./paper_code/data/MajorRivers/MajorRivers.shp")
siletz <- st_read("./data/siletzshapefile/siletz2.shp")
salmon <- st_read("./data/Salmon/salmon.shp")
butte <- st_read("./data/Butte Creek/butte_creek2.shp")
battle <- st_read("./data/Battle Creek/battle_creek2.shp")
deer <- st_read("./data/Deer Creek/deer_creek2.shp")
# this is how we saved them all
# save(cali_rivers, siletz, salmon, butte, battle, deer, 
#     file = "stored_results/100/rivers.rda", 
#     compress = "xz")
```

```{r}
load("stored_results/100/rivers.rda")
```

## Transform

```{r}
#reproject all files to basemap coords
cali_rivers_repro <- st_transform(cali_rivers, "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")
siletz_repro <- st_transform(siletz, "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")
salmon_repro <- st_transform(salmon, "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")
battle_repro <- st_transform(battle, "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")
butte_repro <- st_transform(butte, "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")
deer_repro <- st_transform(deer, "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")
```


## Crop
```{r}
# important to put them in this order and named like this
domain <- c(
  xmin = -130, 
  xmax = -115,
  ymin = 35.5,
  ymax = 45.5
)
nat.crop <- crop(nat.earth, y=extent(domain))
state.subset <- st_crop(state_prov, domain)
coastline_cropped <- st_crop(coastline, domain)
cali2plot <- st_crop(st_zm(cali_rivers_repro), domain)
rosa_rivers <- cali2plot %>%
  filter(GNIS_Name %in% c("Sacramento River","Russian River", "Eel River", "Trinity River", 
                        "Klamath River", "San Joaquin River", "Feather River"))
```


# Grab data for sampling sites

```{r}
fish_sites <- read_csv("./data/100-RoSA_sample_sites.csv")
```

# Make the base map
```{r base-map-with-sampling-sites}
base_map <- ggplot() +
  ggspatial::layer_spatial(nat.crop) +
  geom_sf(data = state.subset, color = 'gray30', fill = NA) +
  geom_sf(data = coastline_cropped, color = "gray30", fill = NA) +
  geom_sf(data = rosa_rivers, colour = "blue", size=0.5) +
  geom_sf(data = siletz_repro, colour = "blue", size=0.5) +
  geom_sf(data = salmon_repro, colour = "blue", size=0.5) +
  geom_sf(data = battle_repro, colour = "blue", size=0.5) +
  geom_sf(data = butte_repro, colour = "blue", size=0.5) +
  geom_sf(data = deer_repro, colour = "blue", size=0.5) +
  geom_text(data = fish_sites,aes(x = name_long, y = name_lat,label = pop, family = "serif", fontface = font_type), size=3.5) +
  geom_segment(data = fish_sites%>%filter(!pop%in%c("San Joaquin","River (F)")), mapping = aes(x = line_long, xend = longitude, y = name_lat-0.05, yend = latitude), size = 0.5) +
    geom_segment(data = fish_sites%>%filter(pop=="San Joaquin"), mapping = aes(x = line_long, xend = longitude, y = name_lat-0.55, yend = latitude), size = 0.5) +
  geom_point(data = fish_sites, mapping = aes(x = longitude, y = latitude),fill=fish_sites$color2, colour=fish_sites$color2) +
  scale_x_continuous(expand = c(0, 0)) +
  xlab("Longitude")+
  scale_y_continuous("Latitude",expand = c(0, 0)) +
  coord_sf(xlim = domain[1:2],  ylim = domain[3:4]) +
  scalebar(x.min = domain[1], x.max = domain[2], y.min = domain[3], y.max = domain[4], 
           location = "bottomright", model = "WGS84", dist = 250, anchor = c(x = -123, y = 36), st.size = 2.5, transform = TRUE, dist_unit = "km") +
  theme_bw()+
  theme(panel.border = element_rect(colour="black", size=1),
        axis.text.x = element_text(size=8, family = "serif", angle = 35, hjust = 1),
        axis.text.y = element_text(size=8, family = "serif"),
        axis.title.y = element_text(family = "serif",size=10),
        axis.title.x = element_text(family = "serif",vjust=2,size=10),
        plot.margin = margin(0,0.1,0,0.15, "cm"),
        legend.position = "none") 

#base_map
```

## add north arrow to map.
```{r adding-arrow-to-base-map}
arrow_map <- base_map +
  north(x.min = domain[-1], x.max = domain[2], y.min = domain[3], y.max = domain[4], location = "bottomright", anchor = c(x = -120.25, y = 36.5))

#arrow_map
```

## Now, work on the world-scale map with the inset:
```{r inset-map}
wrld <- map_data("state")
domain_df <- data_frame(point = 1:length(domain), long = rep(domain[1:2], each = 2), lat = c(domain[3:4], rev(domain[3:4])))
inset_world <- ggplot() + 
  geom_path(data = wrld, aes(x=long, y=lat, group=group), colour="black", size = 0.1) +
  geom_polygon(data = domain_df, mapping = aes(x = long, y = lat), colour = "red", fill = "red", alpha = 0.3) +
  coord_map("ortho", orientation=c(41, -132, 0)) +
  theme_bw() +
  labs(x = NULL, y = NULL) +
  theme(axis.title= element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        plot.margin = unit(c(0,0,-1,-1), "mm"))
#inset_world
```

## Use cowplot to make the inset appear on the bigger map

```{r cowplot-to-make-inset}
final_map <- ggdraw() +
  draw_plot(arrow_map) +
  draw_plot(inset_world, x=0.7, y=0.725, width=0.25, height = 0.2) 
   
ggsave(final_map, filename = "./outputs/100/RoSA_figure1_map_with_inset.pdf", width = 4.75, height=4.25)
# and then plot it in the notebook (albeit at different size) too
final_map
```



# Session Info

```{r}
sessioninfo::session_info()
```

# Running Time

Running the code and rendering this notebook required approximately this much time
on a Mac laptop of middling speed:
```{r}
Sys.time() - start_time
```