---
title: "102.1-klamath-estuary-partial-correlation"
author: "Neil Thompson"
date: "Last Updated: `r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
start_time <- Sys.time()
```

# Partial correlation analyses of maturation status and adiposity with fixed effects

```{r}
library(tidyverse)
library(ppcor)
```

1. load RoSA data for estuary samples
```{r load-data}
rosa_data <- read_rds("./data/101-RoSA-klamath-estuary-samples.rds")

klaten_nomiss <- rosa_data %>%
  distinct(Indiv, .keep_all = T) %>%
  filter(!str_detect(rosa, "\\?"))
```

There are `r nrow(klaten_nomiss)` at-entry estuary samples with 100% complete genotypes. There are `r nrow(rosa_data)` Klamath at-entry estuary samples.

2. load metadata for estuary samples. Join RoSA data to the metadata.
```{r metadata-pull}
estuary_meta <- read_csv("./data/102-Klamath-entryDATA_withNMFSID.csv")
rosa_meta <- left_join(estuary_meta, rosa_data, by = c("NMFS ID" = "NMFS_DNA_ID")) %>%
  dplyr::select(-ID) %>%
  rename(NMFS_DNA_ID = "NMFS ID") %>%
  dplyr::select(NMFS_DNA_ID, rosa, everything())
```

Make dataframe to calculate partial correlations from.
```{r df-for-partial-correlations}
rosa_stats <- rosa_meta %>%
  filter(
    !str_detect(rosa, "\\?"),
    rosa %in% c("EEEEEEEE", "HHHHHHHH", "LLLLLLLL")
  ) %>% # remove samples with missing data and recombinant genotypes
  mutate(
    log_jday = log(julian),
    rosa_f = as.numeric(factor(rosa, levels = c("EEEEEEEE", "HHHHHHHH", "LLLLLLLL"))),
    year_f = as.numeric(factor(year, levels = c("2009", "2010"))),
    sex_f = as.numeric(factor(sex, levels = c("f", "m")))
  )

str(rosa_stats)
```


partial correlation between fatness and sampling date controlling for RoSA, year and sex.
```{r fatness-partial-correlation-julianday}
pcor.test(rosa_stats$drywet, rosa_stats$julian, rosa_stats[, c("rosa_f", "year_f", "sex_f")]) #-0.255
```

partial correlation between GSI and sampling date controlling for RoSA, year and sex.
```{r, gonadsi-partial-correlation-julianday}
pcor.test(rosa_stats$gsi, rosa_stats$julian, rosa_stats[, c("rosa_f", "year_f", "sex_f")]) # 0.533
```


Partial correlation analysis between fatness and RoSA genotype, controlling for sex, and year as fixed effects.

```{r fatness-partial-correlation-rosa}
pcor.test(rosa_stats$drywet, rosa_stats$rosa_f, rosa_stats[, c("julian", "year_f", "sex_f")]) # 0.079
```

Partial correlation analysis between GSI and RoSA genotype, controlling for sex, and year as fixed effects.
```{r gonadsi-partial-correlation-rosa}
pcor.test(rosa_stats$gsi, rosa_stats$rosa_f, rosa_stats[, c("julian", "year_f", "sex_f")]) #-0.096
```

Little correlation between RoSA genotype and adiposity or maturation status. This is a less formal test that RoSA has little influence on the physiological traits thought to be central to the spring-run and fall-run ecotype differentiation.



Partial correlations using log(julian_day) instead of julian day. I doubt this will change much of anything, but log(julian_day) is the predictor variable we used in the mixed effects model selection analyses.
```{r fatness-partial-correlation-logjulian}
pcor.test(rosa_stats$drywet, rosa_stats$log_jday, rosa_stats[, c("rosa_f", "year_f", "sex_f")]) #-0.242

pcor.test(rosa_stats$gsi, rosa_stats$log_jday, rosa_stats[, c("rosa_f", "year_f", "sex_f")]) # 0.476
```
As suspected, those partial correlations are not very different from the partial correlations using julian day.

# Session Info

```{r}
sessioninfo::session_info()
```

# Running Time

Running the code and rendering this notebook required approximately this much time
on a Mac laptop of middling speed:
```{r}
Sys.time() - start_time
```
