---
title: "RoSA manuscript fig4 gonadsi fatness entry timing"
author: "Neil Thompson"
date: "Last Updated: `r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
start_time <- Sys.time()
```

# Code to recreate fig 4 in the RoSA ms

This is going to be a multi panel figure. The estuary sampling date in the left panel (full height) with the gonadsi and fatness data in the right panel.

```{r}
library(tidyverse)
library(cowplot)
dir.create("outputs/105", recursive = TRUE, showWarnings = FALSE)
```

```{r load-data}
rosa_data <- read_rds("./data/101-RoSA-klamath-estuary-samples.rds")

estuary_meta <- read_csv("./data/102-Klamath-entryDATA_withNMFSID.csv")

rosa_meta <- left_join(estuary_meta, rosa_data, by = c("NMFS ID" = "NMFS_DNA_ID")) %>%
  dplyr::select(-ID) %>%
  rename(NMFS_DNA_ID = "NMFS ID") %>%
  dplyr::select(NMFS_DNA_ID, rosa, everything())
```

Remove all samples with ANY missing data. And the recombinant genotypes.

```{r}
rosa_plot <- rosa_meta %>%
  filter(
    !str_detect(rosa, "\\?"),
    rosa %in% c("EEEEEEEE", "HHHHHHHH", "LLLLLLLL")
  )
```


```{r gonadsi-all-haplotypes-figure}
sex_names <- list(
  "f" = "Female",
  "m" = "Male",
  "EEEEEEEE" = "EE",
  "HHHHHHHH" = "EL",
  "LLLLLLLL" = "LL"
)

sex_labeller <- function(variable, value) {
  return(sex_names[value])
}

rosa_meta %>%
  filter(!str_detect(rosa, "\\?")) %>%
  mutate(comp_geno = factor(rosa, levels = c("EEEEEEEE", "HHHHHHHH", "LLLLLLLL", "HHHEEEEE", "LLLLHHLL"))) %>%
  ggplot(., aes(x = julian, y = gsi, fill = comp_geno, colour = comp_geno)) +
  geom_point() +
  xlab(label = "Estuary sampling date") +
  ylab(label = "Gonadosomatic Index") +
  scale_fill_discrete(name = "Genotype") +
  scale_colour_discrete(name = "Genotype") +
  theme_bw() +
  scale_x_continuous(breaks = c(121, 152, 182, 213, 244, 274, 305), labels = c(
    "May-01", "June-01",
    "July-01", "Aug-01", "Sept-01", "Oct-01", "Nov-01"
  )) +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14), axis.text.x = element_text(angle = 45, hjust = 1), strip.background = element_blank()
  ) +
  facet_grid(. ~ sex, labeller = sex_labeller)
```

# recreate plot from above with recombinant genotypes.
```{r gonadsi-non-recombinant-figure}
gonadsi_figure <- rosa_plot %>%
  mutate(comp_geno = factor(rosa, levels = c("EEEEEEEE", "HHHHHHHH", "LLLLLLLL"))) %>%
  ggplot(., aes(x = julian, y = gsi, fill = comp_geno, colour = comp_geno)) +
  geom_point(alpha = 0.75) +
  xlab(label = "Estuary sampling date") +
  ylab(label = "Gonadosomatic Index") +
  scale_fill_discrete(name = "Genotype") +
  scale_colour_manual(name = "Genotype", values = c("gold", "tan2", "blue"), labels = c("EE", "EL", "LL")) +
  theme_bw() +
  scale_x_continuous(limits = c(121, 305), breaks = c(121, 152, 182, 213, 244, 274, 305), labels = c("", "", "", "", "", "", "")) +
  scale_y_continuous(breaks = c(0, 0.05, 0.10, 0.15), labels = c(0, 0.05, 0.10, 0.15), limits = c(0, 0.16), position = "right") +
  theme(
    axis.text = element_text(size = 12, family = "serif"),
    axis.title = element_text(size = 12, family = "serif"), axis.text.x = element_text(angle = 45, hjust = 1, family = "serif"),
    strip.background = element_blank(), legend.position = "none",
    panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
    axis.title.x = element_blank(), strip.text.x = element_text(size = 14, family = "serif"), plot.margin = unit(c(0, 0, 0, 0), "lines")
  ) +
  facet_grid(. ~ sex, labeller = sex_labeller)

gonadsi_figure
```

Now lets do that with the fatness data
```{r fatness-non-recombinant-figure}
fatness_figure <- rosa_plot %>%
  mutate(comp_geno = factor(rosa, levels = c("EEEEEEEE", "HHHHHHHH", "LLLLLLLL"))) %>%
  ggplot(., aes(x = julian, y = drywet, fill = comp_geno, colour = comp_geno)) +
  geom_point(alpha = 0.75) +
  xlab(label = "Estuary sampling date") +
  ylab(label = "Liver NWF") +
  scale_fill_discrete(name = "Genotype") +
  scale_colour_manual(name = "Genotype", values = c("gold", "tan2", "blue"), labels = c("EE", "EL", "LL")) +
  theme_bw() +
  scale_x_continuous(limits = c(121, 305), breaks = c(121, 152, 182, 213, 244, 274, 305), labels = c("1 May", "1 Jun", "1 Jul", "1 Aug", "1 Sep", "1 Oct", "1 Nov")) +
  scale_y_continuous(position = "right") +
  theme(
    axis.text = element_text(size = 12, family = "serif"),
    axis.title = element_text(size = 14, family = "serif"), axis.text.x = element_text(angle = 45, hjust = 1, family = "serif"),
    strip.background = element_blank(), legend.position = "none",
    panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
    strip.text.x = element_blank(), plot.margin = unit(c(-0.5, .2, 0.2, 0.2), "lines")
  ) +
  facet_grid(. ~ sex, labeller = sex_labeller)

fatness_figure
```

Make the estuary date by RoSA genotype figure.
```{r}
entry_figure <- rosa_plot %>%
  mutate(comp_geno = factor(rosa, levels = c("EEEEEEEE", "HHHHHHHH", "LLLLLLLL"))) %>%
  ggplot(., aes(x = julian, fill = comp_geno)) +
  geom_bar(stat = "count", width = 1, position = "dodge") +
  xlab(label = "Estuary sampling date") +
  ylab(label = "Count") +
  scale_fill_manual(
    values = c("gold", "tan2", "blue"),
    breaks = c("EEEEEEEE", "HHHHHHHH", "LLLLLLLL"),
    labels = c("EE", "EL", "LL"),
    name = "RoSA genotype"
  ) +
  theme_bw() +
  background_grid(major = "xy", minor = "none") +
  scale_x_continuous(limits = c(121, 305), breaks = c(121, 152, 182, 213, 244, 274, 305), labels = c("1 May", "1 Jun", "1 Jul", "1 Aug", "1 Sep", "1 Oct", "1 Nov")) +
  scale_y_continuous(limits = c(0, 30)) +
  theme(
    axis.text = element_text(size = 12, family = "serif"),
    axis.title = element_text(size = 14, family = "serif"), axis.text.x = element_text(angle = 45, hjust = 1, family = "serif"),
    strip.background = element_blank(), strip.text.y = element_blank(), legend.position = "none",
    panel.grid.major = element_blank(), panel.grid.minor = element_blank()
  ) +
  facet_grid(comp_geno ~ .)

entry_figure
```

Need to harvest a legend from one of the plots

```{r}
entry_figure_legend <- rosa_plot %>%
  mutate(comp_geno = factor(rosa, levels = c("EEEEEEEE", "HHHHHHHH", "LLLLLLLL"))) %>%
  ggplot(., aes(x = julian, fill = comp_geno)) +
  geom_bar(stat = "count", width = 1, position = "dodge") +
  xlab(label = "Estuary sampling date") +
  ylab(label = "Count") +
  scale_fill_manual(
    values = c("gold", "tan2", "blue"),
    breaks = c("EEEEEEEE", "HHHHHHHH", "LLLLLLLL"),
    labels = c("EE", "EL", "LL"),
    name = "RoSA genotype"
  ) +
  theme_bw() +
  background_grid(major = "xy", minor = "none") +
  scale_x_continuous(limits = c(121, 305), breaks = c(121, 152, 182, 213, 244, 274, 305), labels = c("1 May", "1 Jun", "1 Jul", "1 Aug", "1 Sep", "1 Oct", "1 Nov")) +
  scale_y_continuous(limits = c(0, 30)) +
  theme(
    axis.text = element_text(size = 12, family = "serif"),
    axis.title = element_text(size = 14, family = "serif"), axis.text.x = element_text(angle = 45, hjust = 1, family = "serif"),
    strip.background = element_blank(), strip.text.y = element_blank(),
    panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
    legend.text = element_text(size = 10, family = "serif"),
    legend.title = element_text(size = 10, family = "serif"), legend.key.size = unit(1, "line")
  ) +
  facet_grid(comp_geno ~ .)

legend <- get_legend(entry_figure_legend + theme(legend.position = "top", legend.direction = "horizontal"))
```


## cowplot em together
```{r cowplot-together}
library(cowplot)
pheno_traits <- plot_grid(gonadsi_figure, fatness_figure, labels = c("B", "C"), ncol = 1, align = "v", label_y = c(0.85, 1.05))
entry_trait <- plot_grid(entry_figure, ncol = 1, labels = c("A"))
all_panel <- plot_grid(entry_trait, pheno_traits, ncol = 2, align = "hv", rel_widths = c(0.75, 1))

plot_grid(legend, all_panel, ncol = 1, rel_heights = c(0.05, 1))
ggsave("./outputs/105/RoSA_figure4_multipanel_estuary_gonadsi_fatness.pdf", width = 7.5, height = 4)
ggsave("./outputs/105/RoSA_figure4_multipanel_estuary_gonadsi_fatness.png", width = 7.5, height = 4)
```


Make a supplemental entry date figure showing all recombinants

```{r entry-date-supplement}
rosa_meta %>%
  filter(!str_detect(rosa, "\\?")) %>%
  mutate(comp_geno = factor(rosa, levels = c("EEEEEEEE", "HHHHHHHH", "LLLLLLLL", "HHHEEEEE", "LLLLHHLL"))) %>%
  ggplot(., aes(x = julian, fill = comp_geno)) +
  geom_bar(stat = "count", width = 1, position = "dodge") +
  xlab(label = "Estuary sampling date") +
  ylab(label = "Count") +
  scale_fill_manual(
    values = c("gold", "tan2", "blue", "black", "black"),
    breaks = c("EEEEEEEE", "HHHHHHHH", "LLLLLLLL", "HHHEEEEE", "LLLLHHLL"),
    name = "RoSA genotype"
  ) +
  theme_bw() +
  background_grid(major = "xy", minor = "none") +
  scale_x_continuous(limits = c(121, 305), breaks = c(121, 152, 182, 213, 244, 274, 305), labels = c("1 May", "1 Jun", "1 Jul", "1 Aug", "1 Sep", "1 Oct", "1 Nov")) +
  scale_y_continuous(limits = c(0, 20), breaks = seq(0, 20, 5), labels = seq(0, 20, 5)) +
  theme(
    axis.text = element_text(size = 12, family = "serif"),
    axis.title = element_text(size = 14, family = "serif"), axis.text.x = element_text(angle = 45, hjust = 1, family = "serif"),
    strip.background = element_blank(), strip.text.y = element_text(size = 8),
    panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.position = "none"
  ) +
  facet_grid(comp_geno ~ year) +
  ggtitle("all RoSA genotypes by estuary sampling date")
```


Make a supplemental gonadsi figure showing all recombinants
```{r}
all_nads_09 <- rosa_meta %>%
  filter(
    !str_detect(rosa, "\\?"),
    year == "2009"
  ) %>%
  mutate(
    hapstr = factor(rosa, levels = c("EEEEEEEE", "HHHHHHHH", "LLLLLLLL", "HHHEEEEE", "LLLLHHLL")),
    sex = str_replace(sex, "f", "Female"),
    sex = ifelse(sex == "m", "Male", sex)
  ) %>%
  ggplot(., aes(x = julian, y = gsi, fill = hapstr, colour = hapstr)) +
  geom_point() +
  xlab(label = "Estuary sampling date") +
  ylab(label = "Gonadosomatic Index") +
  scale_color_manual(
    values = c("gold", "tan2", "blue", "black", "black"),
    breaks = c("EEEEEEEE", "HHHHHHHH", "LLLLLLLL", "HHHEEEEE", "LLLLHHLL"),
    name = "RoSA genotype"
  ) +
  theme_bw() +
  scale_x_continuous(limits = c(121, 305), breaks = c(121, 152, 182, 213, 244, 274, 305), labels = c("1 May", "1 Jun", "1 Jul", "1 Aug", "1 Sep", "1 Oct", "1 Nov")) +
  scale_y_continuous(breaks = c(0, 0.05, 0.10, 0.15), labels = c(0, 0.05, 0.10, 0.15), limits = c(0, 0.16)) +
  theme(
    axis.text = element_text(size = 12, family = "serif"),
    axis.title = element_text(size = 14, family = "serif"), axis.text.x = element_text(angle = 45, hjust = 1, family = "serif"),
    strip.background = element_blank(), strip.text.y = element_text(size = 8),
    panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.position = "none"
  ) +
  facet_grid(hapstr ~ sex) +
  ggtitle("2009") +
  theme(plot.title = element_text(hjust = 0.5))



all_nads_10 <- rosa_meta %>%
  filter(
    !str_detect(rosa, "\\?"),
    year == "2010"
  ) %>%
  mutate(
    hapstr = factor(rosa, levels = c("EEEEEEEE", "HHHHHHHH", "LLLLLLLL", "HHHEEEEE", "LLLLHHLL")),
    sex = str_replace(sex, "f", "Female"),
    sex = ifelse(sex == "m", "Male", sex)
  ) %>%
  ggplot(., aes(x = julian, y = gsi, fill = hapstr, colour = hapstr)) +
  geom_point() +
  xlab(label = "Estuary sampling date") +
  ylab(label = "Gonadosomatic Index") +
  scale_color_manual(
    values = c("gold", "tan2", "blue"),
    breaks = c("EEEEEEEE", "HHHHHHHH", "LLLLLLLL"),
    name = "RoSA genotype"
  ) +
  theme_bw() +
  scale_x_continuous(limits = c(121, 305), breaks = c(121, 152, 182, 213, 244, 274, 305), labels = c("1 May", "1 Jun", "1 Jul", "1 Aug", "1 Sep", "1 Oct", "1 Nov")) +
  scale_y_continuous(breaks = c(0, 0.05, 0.10, 0.15), labels = c(0, 0.05, 0.10, 0.15), limits = c(0, 0.16)) +
  theme(
    axis.text = element_text(size = 12, family = "serif"),
    axis.title = element_text(size = 14, family = "serif"), axis.text.x = element_text(angle = 45, hjust = 1, family = "serif"),
    strip.background = element_blank(), strip.text.y = element_text(size = 8),
    panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.position = "none"
  ) +
  facet_grid(hapstr ~ sex) +
  ggtitle("2010") +
  theme(plot.title = element_text(hjust = 0.5))

plot_grid(all_nads_09, all_nads_10)

# ggsave("./outputs/RoSA_klamath_estuary_gonadsi_supplemental_all_haplos_byYear.pdf", width=11, height=8)
# ggsave("./outputs/RoSA_klamath_estuary_gonadsi_supplemental_all_haplos_byYear.png", width=11, height=8)
```

now fatness supplemental
```{r}
rosa_meta %>%
  filter(!str_detect(rosa, "\\?")) %>%
  mutate(
    hapstr = factor(rosa, levels = c("EEEEEEEE", "HHHHHHHH", "LLLLLLLL", "HHHEEEEE", "LLLLHHLL")),
    sex = str_replace(sex, "f", "Female"),
    sex = ifelse(sex == "m", "Male", sex)
  ) %>%
  ggplot(., aes(x = julian, y = drywet, colour = hapstr)) +
  geom_point() +
  xlab(label = "Estuary sampling date") +
  ylab(label = "Liver NWF") +
  scale_color_manual(
    values = c("gold", "tan2", "blue", "black", "red"),
    breaks = c("EEEEEEEE", "HHHHHHHH", "LLLLLLLL", "HHHEEEEE", "LLLLHHLL"),
    name = "RoSA genotype"
  ) +
  theme_bw() +
  scale_x_continuous(limits = c(121, 305), breaks = c(121, 152, 182, 213, 244, 274, 305), labels = c("1 May", "1 Jun", "1 Jul", "1 Aug", "1 Sep", "1 Oct", "1 Nov")) +
  theme(
    axis.text = element_text(size = 12, family = "serif"),
    axis.title = element_text(size = 14, family = "serif"), axis.text.x = element_text(angle = 45, hjust = 1, family = "serif"),
    strip.background = element_blank(), # legend.position = "none",
    panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
    plot.margin = unit(c(0, .2, 0.2, 0.2), "lines")
  ) +
  facet_grid(year ~ sex)
ggsave("./outputs/105/RoSA_klamath_estuary_fatness_supplemental_all_haplos_byYear.pdf", width = 6, height = 4)
ggsave("./outputs/105/RoSA_klamath_estuary_fatness_supplemental_all_haplos_byYear.png", width = 6, height = 4)
```


# Session Info

```{r}
sessioninfo::session_info()
```

# Running Time

Running the code and rendering this notebook required approximately this much time
on a Mac laptop of middling speed:
```{r}
Sys.time() - start_time
```

