---
title: "107-Population-genetics-survey-rosa"
author: "Neil Thompson"
date: "Last Updated: `r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,warning=FALSE)
start_time <- Sys.time()
```

```{r}
library(tidyverse)
library(kableExtra)
library(knitr)

dir.create("outputs/107", showWarnings = FALSE, recursive = TRUE)
```


NT note: the rmd used to produce rosa_nodups is called "Thompson-etal-RoSA-amplicon-data.rmd" and is in the paper_code folder on NT's laptop.

1. load rosa data that has been prefiltered to remove duplicates and missing data.
```{r, load data}
rosa_nodups <- read_rds("./data/107-RoSA-popgen-survey-data.rds")
```

First let's look at how many samples have recombinant genotypes.

```{r, calculating-number-of-recombinant-RoSA-genotypes}
rosa_nodups %>% 
  filter(!str_detect(rosa,"\\?")) %>% 
  mutate(is_recomb = !rosa %in% c("HHHHHHHH","LLLLLLLL","EEEEEEEE")) %>%
  group_by(is_recomb) %>%
  summarise(n_type = n()) %>% 
  kable("html") %>%
  kable_styling("striped", full_width = FALSE)

n_recomb <- rosa_nodups %>% 
    filter(!str_detect(rosa,"\\?")) %>% 
  mutate(is_recomb = !rosa %in% c("HHHHHHHH","LLLLLLLL","EEEEEEEE")) %>%
  group_by(is_recomb) %>%
  summarise(n_type = n()) %>%
 .[2,"n_type"]

n_nonrecomb <- rosa_nodups %>% 
    filter(!str_detect(rosa,"\\?")) %>% 
  mutate(is_recomb = !rosa %in% c("HHHHHHHH","LLLLLLLL","EEEEEEEE")) %>%
  group_by(is_recomb) %>%
  summarise(n_type = n()) %>% .[1,"n_type"]
```

There are `r n_recomb` recombinants which makes a `r round(n_recomb/sum(n_recomb,n_nonrecomb),3)` frequency of recombinant genotypes.

# Frequency table of non-recombinant haplotypes

Create a frequency table of non-recombinant genotypes in the populations we want for the paper.
```{r, frequency-table-of-RoSA-Table-S7}
paper_pops <- c("Siletz River spring","Siletz River fall","Iron Gate hatchery fall","Salmon River spring","Salmon River fall","Salmon River unknown", "Trinity River hatchery spring", "Trinity River hatchery fall", "Trinity River unknown","Eel River fall","Russian River fall","Sacramento River winter","Coleman hatchery late-fall","Butte Creek spring","Butte Creek fall","Butte Creek unknown","Mill-Deer Creek","Feather River hatchery spring","Feather River hatchery fall","Feather River unknown")

freq_table <- rosa_nodups  %>%
  filter(rosa%in%c("HHHHHHHH","LLLLLLLL","EEEEEEEE")) %>%
  mutate(rosa = factor(rosa, levels = c("EEEEEEEE","HHHHHHHH","LLLLLLLL"))) %>% 
  group_by(pop_name) %>%
  mutate(n_pop = n()) %>%
  group_by(pop_name,rosa) %>%
  mutate(n_rosa = n(),
         freq_rosa = round(n_rosa/n_pop,3)) %>%
  ungroup() %>%
  dplyr::select(pop_name,n_pop,rosa,freq_rosa) %>% 
  distinct(pop_name,rosa, .keep_all=TRUE) %>%
  spread(rosa,freq_rosa) %>%
  arrange(factor(pop_name, levels=paper_pops)) %>%
  dplyr::select(pop_name,contains("EE"),contains("HH"),contains("LL"),n_pop) %>%
  rename(Collection = pop_name,
         EE = EEEEEEEE,
         EL = HHHHHHHH,
         LL = LLLLLLLL,
         n = n_pop) %>%
  mutate(Ecotype = str_extract(Collection,"([spring]{6}|[fall]{4}|[unknown]{7}|[winter]{6}|[late]{4}[-]{1}[fall]{4})"),
         Ecotype = gsub("unknown","mix",Ecotype),
         Ecotype = gsub("spring","spring-run",Ecotype),
         Ecotype = gsub("fall","fall-run",Ecotype),
         Ecotype = gsub("winter","winter-run",Ecotype),
         Ecotype = gsub("late-fall","late fall-run",Ecotype),
         Ecotype = ifelse(Ecotype=="late fall-run-run", "late fall-run",Ecotype)) %>%
  dplyr::select(Collection,Ecotype, everything()) %>%
  replace_na(list(Ecotype="mix")) %>%
  mutate(Collection = recode(
    Collection, 
    `Coleman hatchery late-fall` = "Battle Creek late-fall",
    `Iron Gate hatchery fall` = "Klamath River fall"
    ))


freq_table %>% 
  replace_na(list(EE="", EL="", LL=""))


##write out frequency table
freq_table %>% 
  replace_na(list(EE = "0.000", EL = "0.000", LL = "0.000")) %>%
  write_csv("outputs/107/RoSA_frequencies_tableS7.csv")
```

# Frequency table including the within-RoSA recombinants

Create a table of all RoSA genotypes including recombinants.
```{r, RoSA-all-genos-count-table}
rosa_hap_order <- rosa_nodups %>%
  filter(!str_detect(rosa,"\\?")) %>%
  distinct(rosa) %>%
  mutate(n_e = str_count(rosa,"E"),
         n_h = str_count(rosa,"H"),
         n_l = str_count(rosa,"L"))%>%
  arrange(desc(n_e), desc(n_h), desc(n_l)) %>%
  pull(rosa)

count_table <- rosa_nodups %>%
  filter(!str_detect(rosa,"\\?")) %>%
 mutate(rosa_f = factor(rosa, levels = rosa_hap_order)) %>% 
  group_by(pop_name) %>%
  mutate(n_pop = n()) %>%
  group_by(pop_name,rosa) %>%
  mutate(n_rosa = n()) %>%
  ungroup() %>%
  dplyr::select(pop_name,n_pop,rosa_f,n_rosa) %>%
  distinct(pop_name,rosa_f, .keep_all=TRUE) %>% 
  spread(rosa_f,n_rosa) %>%
  arrange(factor(pop_name, levels=paper_pops)) %>%
  dplyr::select(pop_name,rosa_hap_order,n_pop) %>%
  rename(Collection=pop_name,
         n = n_pop) %>%
  mutate(Collection = recode(
    Collection, 
    `Coleman hatchery late-fall` = "Battle Creek late-fall",
    `Iron Gate hatchery fall` = "Klamath River fall"
    ))

# new versions of tidyr don't allow type conversions for NA replacements, so
# deal with it this way
count_table_char <- count_table %>%
  mutate_all(.funs = as.character)

count_table_char[is.na(count_table_char)] <- ""

##write out count table here.
write_csv(count_table_char, "outputs/107/RoSA_counts_all_genotypes_dataS3.csv")
```

# Session Info

```{r}
sessioninfo::session_info()
```

# Running Time

Running the code and rendering this notebook required approximately this much time
on a Mac laptop of middling speed:
```{r}
Sys.time() - start_time
```